Wine Analysis by Mathias Wainer
========================================================

```{r packages, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
#all libraries here
library(ggplot2)
library(dplyr)
library(scales)
library(reshape2)
library(lubridate)
library(gridExtra)
library(swirl)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wineData <- read.csv('wineQualityReds.csv')
```

This dataset is composed of 1599 observations of different red wines and 11
variables. The dataset was created by: Paulo Cortez (Univ. Minho), 
Antonio Cerdeira, Fernando Almeida, Telmo Matos and Jose Reis. By making this 
data publicly available it allows everyone to explore and come up with original
analysis. The output variable (y) is the quality of wine as determined by wine
someliers. It is the hope of the creators and anyone who explores the data to
come up with conclusions on what variables make the best wines.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
names(wineData)
summary(wineData)
summary(wineData$quality)

round(table(wineData$quality)/ nrow(wineData), digits = 3)

```

We can see above x being all the observations, the 11 variables and 'quality'
as the output variable. We also can look at the summary statistics for each one.
Quality was printed again since it should be our focus to decipher how it is
shaped by every variable.

```{r echo=FALSE, Univariate_histograms}

h1 <- ggplot(aes(x = quality), data = wineData)+
  geom_bar()
h2 <- qplot(x = fixed.acidity, data = wineData)
h3 <- qplot(x = volatile.acidity, data = wineData)
h4 <- qplot(x = citric.acid, data = wineData)
h5 <- qplot(x = residual.sugar, data = wineData)
h6 <- qplot(x = chlorides, data = wineData)
h7 <- qplot(x = free.sulfur.dioxide, data = wineData)
h8 <- qplot(x = total.sulfur.dioxide, data = wineData)
h9 <- qplot(x = density, data = wineData)
h10 <- qplot(x = pH, data = wineData)
h11<- qplot(x = sulphates, data = wineData)
h12<- qplot(x = alcohol, data = wineData)

grid.arrange(h1,h2,h3,h4, ncol = 2)


```

We can see that quality is clustred around scores 5 and 6 (80% of observations).
Both fixed and volatile acidities follow what appears to be a normal 
distribution. Citric.acid does not follow a normal distribution and may even be
bimodal.

```{r}
grid.arrange(h5,h6,h7,h8, ncol = 2)

```

We can see that residual.sugar and chlorides have some extreme right tailed
values. Total.sulfur.dioxide has 2 values near the 300 range but is otherwise
under 100 sulfates (g/dm^3)

```{r}
grid.arrange(h9,h10,h11,h12, ncol = 2)

```

These 4 grapgs appear to be quite normally distributed. Graphing sulphates by
limiting the x axis around 1.25 (g/dm^3) may express the distribution more
accurately.


```{r echo=FALSE, zoomed_histograms}
qplot(x = residual.sugar, data = wineData, binwidth = .1) +
  xlim(0,5)
```

After limiting the x axis for residual.sugar we can see the distribution be 
more normal and center around 2.25.

```{r echo=FALSE, zoomed_histograms2}
qplot(x = chlorides, data = wineData, binwidth = .005) +
  xlim(0,.2)

```
 
After limiting the x axis for chlorides we can see the distribution be 
more normal and center around 0.8 (g/dm^3).
 
 
```{r echo=FALSE, freqpoly_graphs}

ggplot(data = wineData, aes(x = density)) +
 geom_freqpoly()

ggplot(data = wineData, aes(x = citric.acid)) +
 geom_freqpoly()

```

These two line graphs show the normal distribution of density and the non-
normal distribution of citric.acid.

# Univariate Analysis

After creating charts for all the variables in the dataset I was pleasently 
surprised. Most of variable frequencies follow a pretty well defined normal
distribution, examples are volatile.acidity and fixed.acidity.

The variables residual.sugar and chlorides and some other variables (to some
extent) had some extreme right values. Once the x axis were limited to exclude
extreme values the data became more normal. 

Something to explore in the bivariate analysis section is the citric.acid 
variable. It does not appear to have a normal distribution. 

Something to note is that over 81% the wine quality scores are either 5 or 6. 
To get better correlation statistics and figure out what variables affect it
would be beneficial to have a wide range of quality scores.


# Bivariate Plots Section


```{r}
ggplot(aes(x = alcohol, y = density), data = wineData) + geom_point()
cor.test(wineData$alcohol, wineData$density)

ggplot(aes(x = free.sulfur.dioxide, y = total.sulfur.dioxide), 
       data = wineData) + geom_point()
cor.test(wineData$free.sulfur.dioxide, wineData$total.sulfur.dioxide)
```


First I graphed density vs alcohol which should be negatively correlated since
alcohol is less dense than water. A spearson correlation test confirmed it. I
did the same for free.sulfur and total.sulfur. Predictably total sulfur was
positively correlated with free sulfur. 


```{r echo=FALSE, Bivariate_Plots}

ggplot(aes(x = fixed.acidity, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = volatile.acidity, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = citric.acid, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = residual.sugar, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')+
  xlim(0,5)

ggplot(aes(x = chlorides, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(0,.2)

ggplot(aes(x = free.sulfur.dioxide, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = total.sulfur.dioxide, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = density, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = pH, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = sulphates, y = quality), data = wineData) + 
  geom_jitter(alpha = 1/5)+
  geom_line(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')

ggplot(aes(x = alcohol, y = quality), data = wineData, color = 'orange') + 
  geom_jitter(alpha = 1/5)+
  geom_smooth(stat = 'summary', fun.y = mean, color = 'blue') +
  geom_smooth(method = 'lm', color = 'red')


```

After graphing each variable against quality. I added some transparency for
readibility and added a mean line (blue) and a linear regression line (red).
After visual analysis i would like to run pearson correlation for only 4 of the
variables that had large slopes on the red line (fitted line). The 4 largest
slopes were for variables alcohol, sulphates, citric.acid and volatile.acidity.

```{r echo=FALSE, corr_test}
cor.test(wineData$alcohol,wineData$quality)
cor.test(wineData$sulphates,wineData$quality)
cor.test(wineData$citric.acid,wineData$quality)
cor.test(wineData$volatile.acidity,wineData$quality)

```

# Bivariate Analysis

First I ran two scatter plots of variables I expected would be correlated to a 
high degree (alcohol vs density, free.sulphates vs total.sulphates) to confirm
my suspicion.

Next I ran each variable against quality and took note of highly-slopped
means and smoothed lines. I identified citric acid, alcohol, volatile.acidity,
and sulphates as possible candidates to have a large regression. 

After computing the correlation if the 4 identified variables only citric acid
had a correlation score between (-0.25,0.25) meaning the correlation was
especially weak.

Alcohol had the largest presented correlation to quality which suggests the
comical conclusion that a drunk wine tester is a happy wine tester. However
future causal research must be performed to verify this.



# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
#bins for alcohol consentration 
wineData$alc_concentrate.bucket <- cut(wineData$alcohol, 
                             c(8,9.5,11,12.5,14))
wineData$alc_concentrate2.bucket <- cut(wineData$alcohol, 
                             c(8,11,14))


#plotting quality by volatile.acidity with alcohol bins
ggplot(aes(x=volatile.acidity, y= quality),
       data = subset(wineData, !is.na(alc_concentrate.bucket))) +
  geom_point(aes(color = alc_concentrate.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm', color = 'red')+
  scale_color_brewer(type='seq')+
  theme_dark()


#plotting quality by citric.acid with alcohol bins
ggplot(aes(x=citric.acid, y= quality),
       data = subset(wineData, !is.na(alc_concentrate2.bucket))) +
  geom_point(aes(color = alc_concentrate2.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm', color = 'red') +
  ylim(4.4,8)+
  theme_dark()

#plotting quality by sulphates with alcohol bins
ggplot(aes(x=sulphates, y= quality),
       data = subset(wineData, !is.na(alc_concentrate.bucket))) +
  geom_point(aes(color = alc_concentrate.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(.3,1.25)+
  scale_color_brewer(type='seq')+
  theme_dark()


```

The graph volatile.acidity vs quality by alcohol buckets appears to have a trend
in which the lower the acidity the better the quality. We can see an 
overwhelming number of blue and purple in the top right of the plot. 
purple

Citric.acid vs quality by alcohol buckets appears to have a positive trend.
However it appears (at least visually) that alcohol content has a larger effect
on quality than citric acid.

The sulphates vs quality by alcohol buckets has many 7 and 8 rated quality wines
in the middle of the x range (around .75) 

```{r echo=FALSE, message=FALSE, model}

firstmodelfit <- glm(quality ~ alcohol+citric.acid+sulphates+volatile.acidity,data = wineData)
summary(firstmodelfit)

secondmodelfit <- glm(quality ~ alcohol+sulphates+volatile.acidity,data = wineData)
summary(secondmodelfit)


obs250 <- wineData[250,]
qual250 <- wineData$quality[250]
obs500 <- wineData[500,]
qual500<- wineData$quality[500]

modelEstimate250 = predict(secondmodelfit, newdata = obs250,
                        interval = 'prediction', level = .95)

modelEstimate500 = predict(secondmodelfit, newdata = obs500,
                        interval = 'prediction', level = .95)

totalEstimate = predict(secondmodelfit, newdata = wineData,
                        interval = 'prediction', level = .95)

qual250
modelEstimate250
qual500
modelEstimate500

summary(wineData$quality)
summary(totalEstimate)



```

After visually inspecting all the graphs I attempted to fit quality to be
dependent on citric.acid, sulphates, volatile.acidity and alcohol. The results
had very poor confidence intervals. The best performing parameters was alchol,
sulphates and volatile.acidity. Running that fitted model was much better
as most variables had a 99.9% confidence interval.

I then pulled the quality score for observation 250, which was six and ran it 
against the model which predicted 5.52 (rounded up thats 6). Obervation 500 had 
a quality score of 6 and the fitted model predicted 5.86 which is pretty close. 


# Multivariate Analysis

After separating alcohol content into four buckets and running scatter plots 
on the other three identified variables several trends became apparent. Alcohol
had the largest effect on quality of content. Very few highly rated wines were
in the two lower alcohol groups. Additionally because of interaction between 
volatile.acids and citric.acid one of the two had to be dropped from the fitted 
model.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
ggplot(aes(x = alcohol, y = quality), data = wineData) + 
  geom_jitter(alpha = .4, color = 'blue')+
  geom_smooth(method = 'lm', color = 'red')+
  xlab('alcohol percentage (% by volume)')+
  ylab('quality score')+
  ggtitle('Alcohol vs Quality')
```

### Description One
  I chose this graph because it illustrates the highest single factor 
  correlation to a wine's quality score. The wines with higher alcohol appear to
  on average to have higher scores. The red fitted line confirms our visual 
  observation.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x=citric.acid, y= quality),
       data = subset(wineData, !is.na(alc_concentrate2.bucket))) +
  geom_point(aes(color = alc_concentrate2.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm', color = 'red') +
  ylim(4.4,8)+
  xlab('citric acid (g/dm^3)')+
  ylab('quality score')+
  ggtitle('Citric acid vs Quality by alcohol')

```

### Description Two
  This graph plots citric acid vs quality score. It is interesting that there
  are essentially no observations above the fitted line that have an alcohol
  content in the lower alcohol bucket. At the same time we can see that a high 
  citric acid content may propel some wines in the lower alcohol category to 7
  and 8 ratings. 

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x=sulphates, y= quality),
       data = subset(wineData, !is.na(alc_concentrate.bucket))) +
  geom_point(aes(color = alc_concentrate.bucket),
            stat = 'summary', fun.y = mean) +
  geom_smooth(method = 'lm', color = 'red') +
  xlim(.3,1.25)+
  ylim(4.5,8)+
   ggtitle('Sulphates vs Quality by alcohol')+
  scale_color_brewer(type='seq')+
  xlab('sulphates (g/dm^3)')+
  theme_dark()
```

### Description Three

I chose the sulphates vs quality plot because it displays how values further
from the fitted mean can accentuate its slope. There are many more wines with
the two lower levels of alcohol content than the two higher ones. Additionaly
the few high quality wines tend to have a higher alcohol content, either due to 
causation or due to sampling. This fact drives the fitted mean line to a more
pronounced slope.
------

# Reflection

During the initial single variable analysis I was pleasently
surprised how normal all the distributions were. It makes sense that most
variables would be normal since wines are produced in a similar
fashion. The biggest struggle during the analysis was that the output variable
'quality' was discrete and only had 5 possible outcomes (3 to 8 range). This
fact makes graphing it more difficult as you don't want too much overlapping
observations clustered around each quality score.

Future analysis of this dataset should focus on developing and testing various
predictive models. These models can then be tested against new observations to
test its efficacy. I read online that it is possible to randomly use only half
a data set to create a model and then have it tested against the remaining half. 
It would be interesting to do this procedure and statistically evaluate it. 

